Hector Permafrost

### 2020-01-23 analysis/paper/paper.Rmd
Update permafrost submodel documentation
Now, it reflects what Hector actually does.

```{r setup, echo = FALSE}
library(drake)
library(knitr)
library(here)

@@

library(hector.permafrost.emit)

@@

The general principle is that permafrost constitutes an additional reserve of soil carbon that, because it is frozen, is inaccessible to microbes.
As permafrost thaws and some fraction of this carbon becomes accessible to microbes, it is transferred from the frozen permafrost pool to the standard soil C pool, where it is decomposed by Hector's standard decomposition routine.
Let $C[t]_{pf}$ be the permafrost C pool and $C[t]_{s}$ be the soil C pool at time $t$ (both in units of Pg C), and let $\Delta C[t]_{pf}$ be the change in the permafrost C pool at time $t$.
The C consequences of permafrost thaw can therefore be represented as:

\[ C[t]_{pf} = C[t-1]_{pf} - \Delta C[t]_{pf} \]

\[ C[t]_{s} = C[t]_{s} + \Delta C[t]_{pf} \]

Let $C[0]_{pf}$ be the initial size of the permafrost C pool (Pg C) and $\Phi[t]$ be the fraction of permafrost remaining (in arbitrary area or volume units) at time $t$.
Assuming a uniform permafrost C density, $\Delta C[t]_{pf}$ can be expressed as:

\[ \Delta C[t]_{pf} = \Phi[t] C[0]_{pf} \]

@@

\[
\Phi[t] = \frac{1}{1 + \exp\left( m T[t]_{air} + b \right)}
\]

where $m$ and $b$ are model parameters.
For globally-averaged permafrost, $m = 0.829$ and $b = 4.078$ most closely reproduce the rate of 0.172 K$^{-1}$ reported by [Kessler (2015)] over the range of 2 to 6 K$^{-1}$.

```{r sigmoid, fig.cap = sigmoid_cap}
sigmoid_cap <- paste(
  "Fraction of permafrost thaw as a function of change in",
  "global annual mean air temperature since pre-industrial (1750)."
)
m <- 0.829
b <- 4.078
kessler_m <- 0.172
kessler_b <- 2 * kessler_m + 1
plot(0, 0, type = "n", xlim = c(0, 10), ylim = c(0, 1),
     xlab = expression(Delta * T[air] ~ (K)),
     ylab = "Frac. permafrost remaining")
curve(1 / (1 + exp(m * x - b)), 0, 10, col = 1, add = TRUE)
abline(a = kessler_b, b = -kessler_m, lty = "dashed")
abline(h = 1, lty = "dotted")
legend("topright", c("Hector", "Kessler (2015)"), lty = c("solid","dashed"),
       bg = "white")
```

@@

```{r permafrost-c, fig.cap = }
permafrost_c_cap <- paste(
  "Effect of permafrost C emissions on scenarios."
)

devtools::load_all("~/Projects/hector_project/hector-permafrost-submodel")
run_rcp <- function(rcp) {
  inidir <- file.path("~", "Projects", "hector_project",
                      "hector-permafrost-submodel", "inst", "input")
  inifile <- normalizePath(file.path(inidir, paste0("hector_rcp", rcp, ".ini")))
  stopifnot(file.exists(inifile))
  hc <- newcore(inifile)
  run(hc)
  dates <- seq(1750, 2100)
  outvars <- c(ATMOSPHERIC_CO2(), GLOBAL_TEMP())
  no_pf <- fetchvars(hc, dates, outvars, scenario = "Original")
  reset(hc)
  setvar(hc, NA, PERMAFROST_C(), 1035, "PgC")
  run(hc)
  yes_pf <- fetchvars(hc, dates, outvars, scenario = "Permafrost")
  return(tibble::as_tibble(dplyr::bind_rows(no_pf, yes_pf)))
}
r26 <- run_rcp("26")
library(ggplot2)
ggplot(r26) +
  aes(x = year, y = value, color = scenario) +
  geom_line() +
  facet_wrap(vars(variable), scales = "free_y")

>>>>

```{r setup, echo = FALSE}
library(drake)
library(knitr)
library(here)
library(dplyr)
library(ggplot2)
library(hector.permafrost.emit)

@@

devtools::load_all("~/Projects/hector_project/hector-permafrost-submodel")

@@

The general principle is that permafrost constitutes an additional reserve of soil carbon that, because it is frozen, is inaccessible to microbes.
As permafrost thaws and some fraction of this carbon becomes accessible to microbes, it is transferred from the frozen permafrost pool to the standard soil C pool, where it is decomposed by Hector's standard decomposition routine.
Let $C_{pf}[t]$ be the permafrost C pool and $C_{s}[t]$ be the soil C pool at time $t$ (both in units of Pg C), and let $\Delta C_{pf}[t]$ be the change in the permafrost C pool at time $t$.
The C consequences of permafrost thaw can therefore be represented as:

\[ C_{pf}[t] = C_{pf}[t-1] - \Delta C_{pf}[t] \]

\[ C_{s}[t] = C_{s}[t] + \Delta C_{pf}[t] \]

Let $C[0]_{pf}$ be the initial size of the permafrost C pool (Pg C) and $\Phi[t]$ be the fraction of permafrost remaining (in arbitrary area or volume units) at time $t$.
Assuming a uniform permafrost C density, $\Delta C_{pf}[t]$ can be expressed as:

\[ \Delta C_{pf}[t] = \Delta \Phi[t] C_{pf}[t-1] \]

@@

\[
\Phi[t] = 1 - \left( 1 + a \exp\left( b * T_{air}[t] \right) \right) ^ c
\]

where $a$, $b$, and $c$ are model parameters.
The change in frozen fraction at a given time step, $\Delta \Phi[t]$, is given by:

\[
\Delta \Phi[t] = \max(\Phi[t] - \Phi[t-1], 0)
\]

In other words, permafrost thaw is permanent -- once it thaws, it does not re-freeze, even if the temperature drops.

For globally-averaged permafrost, $a = 2.371$, $b = -0.676$, and $c = -3.685$ to most closely reproduce the rate of 0.172 K$^{-1}$ reported by [Kessler (2015)][kessler] over the range of 0.8 [^2] to 4 K above the pre-industrial baseline.

[^2]: [Kessler (2015)][kessler] report this as temperature change from year 2000. 0.8 K is the warming since pre-industrial as estimated by the default Hector configuration.

```{r sigmoid, fig.cap = sigmoid_cap}

sigmoid_cap <- paste(
  "Fraction of permafrost thaw as a function of change in",
  "global annual mean air temperature since pre-industrial (1750)."
)
pfa <- 2.371
pfb <- -0.676
pfc <- -3.685
b <- 4.078
kessler_m <- 0.172
kessler_b <- 0.8 * kessler_m + 1
plot(0, 0, type = "n", xlim = c(0, 8), ylim = c(0, 1),
     xlab = expression(Delta * T[air] ~ (K)),
     ylab = "Frac. permafrost remaining")
curve(1 - (1 + pfa * exp(pfb * x)) ^ pfc, 0, 10, col = 1, add = TRUE)
abline(a = kessler_b, b = -kessler_m, lty = "dashed")
abline(h = 1, lty = "dotted")
legend("topright", c("Hector", "Kessler (2015)"), lty = c("solid","dashed"),
       bg = "white")

@@

```{r permafrost-c, fig.cap = permafrost_c_cap}

permafrost_c_cap <- paste(
  "Effect of permafrost C emissions on scenarios."
)

run_rcp <- function(rcp) {
  inidir <- file.path("~", "Projects", "hector_project",
                      "hector-permafrost-submodel", "inst", "input")
  inifile <- normalizePath(file.path(inidir, paste0("hector_rcp", rcp, ".ini")))
  stopifnot(file.exists(inifile))
  hc <- newcore(inifile)
  dates <- seq(1750, 2100)
  run(hc, max(dates))
  outvars <- c(ATMOSPHERIC_CO2(), GLOBAL_TEMP(), PERMAFROST_C(), SOIL_C())
  no_pf <- fetchvars(hc, dates, outvars, scenario = "Original")
  reset(hc)
  setvar(hc, 0, PERMAFROST_C(), 1035, "PgC")
  run(hc, max(dates))
  yes_pf <- fetchvars(hc, dates, outvars, scenario = "Permafrost")
  return(tibble::as_tibble(dplyr::bind_rows(no_pf, yes_pf)))
}

rcps <- c("26", "45", "60", "85")
names(rcps) <- paste0("RCP", rcps)
results <- purrr::map_dfr(rcps, run_rcp, .id = "RCP")

results %>%
  mutate(variable = recode(
    variable,
    Ca = "Atmospheric CO2 (ppm)",
    Tgav = "Global mean temperature anomaly (K)",
    permafrost_c = "Permafrost C pool (PgC)",
    soil_c = "Soil C pool (PgC)"
  )) %>%
  ggplot() +
  aes(x = year, y = value, linetype = scenario, color = RCP) +
  geom_line() +
  facet_wrap(vars(variable), scales = "free_y") +
  theme_bw() +
  theme(axis.title = element_blank())
